Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2022-11-28 manuscript/manuscript.Rmd
addressed some major comments of R2.

As the initial step, simple visualization techniques such scatter plots are frequently used to explore the relationships between variables of interest. Given the number of variables involved in this study, we used principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017]. To detect the effect of training on novices' performance as compared with archaeological samples and handaxe made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test. Compared with other nonparametric tests frequently used in archaeological research for multiple group comparison such as Tukey's test, Games-Howell test does not rely on the assumptions of sample normality, and equal sample sizes and equal variance are not necessary conditions to perform this test. The sample size of each compared group can be as low as 6 [@games1976; @sauder2019]. This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which are available as supplementary materials and can be accessed through the author's Github (<https://github.com/Raylc/Boxgrove-Exp>).

@@

```{r PCcorrelation, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scatter plot showing the correlation between PC1 and PC2 respectively in the groups of Boxgrove (left, n=326), expert (middle, n=10), and novice (right, n=132). The upper left area in each individual plot display statistical reporting from a frequentist perspective, including the student-t test statistics, p-value, Pearson correlation coefficient, confidence interval, and sample size. The lower right area in each individual plot display statistical reporting from a Bayesian perspective, including the natural logarithm of Bayes factor, posterior type and estimate, credible interval, and prior type and value. "}
knitr::include_graphics(here::here("figure/Fig4.png"))
```

@@

Our study suggests that both skill level and mental template have a relatively clear manifestation in different aspects of handaxe morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. In accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (cross-sectional thinning) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove knappers. Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the handaxe into several pieces, serving the purpose of achieving the desired convexity and/or volume. This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013]. Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394], although the validity of differentiating purcussor types (hard hammerstone, soft hammerstone, and antler hammer) based on flake attributes has been challenged by other experimental studies[@driscoll2014]. It should be noted that both our experts and novices frequently used soft hammers in the production of experimental assemblages. In the skill acquisition experiments, novice knappers were explicitly taught to switch to the soft hammer for thinning purposes, but some of them did not follow the instruction during the assessment. On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], and the replicas produced by Bruce Bradley in our expert reference collection did not involve the use of soft hammers.

@@

The negative correlation between the PC1 and PC2 values revealed a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form. Our results (Fig.) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value). In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface. Consequently, it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed. It is possible that such constraints help to explain the convergence of our novice knappers on similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove. Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers. This also provides an alternative explanation to the social transmission of form for the experimental convergence of on pointed forms. In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.

@@

```{r comparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Core 63 before (left) and after knapping(right), showing the minimal morphological change during the process"}
knitr::include_graphics(here::here("figure/Fig7.jpg"))
```

>>>>

We use the statistical programming language R 4.1.1 [@rcoreteam2021] to conduct statistical analyses and data visualization in this study, particularly the R packages "FactoMineR" [@lê2008] and "ggstatsplot" [@patil2021]. As the initial step, simple visualization techniques such scatter plots are frequently used to explore the relationships between variables of interest. Given the number of variables involved in this study, we used principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017]. To detect the effect of training on novices' performance as compared with archaeological samples and handaxe made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test. Compared with other nonparametric tests frequently used in archaeological research for multiple group comparison such as Tukey's test, Games-Howell test does not rely on the assumptions of sample normality, and equal sample sizes and equal variance are not necessary conditions to perform this test. The sample size of each compared group can be as low as 6 [@games1976; @sauder2019]. This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which are available as supplementary materials and can be accessed through the author's Github (<https://github.com/Raylc/Boxgrove-Exp>).

@@

```{r PCcorrelation, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scatter plot showing the correlation between PC1 and PC2 respectively in the groups of Boxgrove (left, n=326), expert (middle, n=10), and novice (right, n=132). The upper left area in each individual plot displays statistical reporting from a frequentist perspective, including the student-t test statistics, p-value, Pearson correlation coefficient, confidence interval, and sample size. The lower right area in each individual plot displays statistical reporting from a Bayesian perspective, including the natural logarithm of Bayes factor, posterior type and estimate, credible interval, and prior type and value. "}
knitr::include_graphics(here::here("figure/Fig4.png"))
```

@@

Our study suggests that both skill level and mental template have a relatively clear manifestation in different aspects of handaxe morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and confirmed the previous finding [-@pargeter2019] that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. In accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (cross-sectional thinning) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove knappers. Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the handaxe into several pieces, serving the purpose of achieving the desired convexity and/or volume. This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013]. Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394], although the validity of differentiating purcussor types (hard hammerstone, soft hammerstone, and antler hammer) based on flake attributes has been challenged by other experimental studies[@driscoll2014]. It should be noted that both our experts and novices frequently used soft hammers in the production of experimental assemblages. In the skill acquisition experiments, novice knappers were explicitly taught to switch to the soft hammer for thinning purposes, but some of them did not follow the instruction during the assessment. On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], and the replicas produced by Bruce Bradley in our expert reference collection did not involve the use of soft hammers.

@@

The negative correlation between the PC1 and PC2 values revealed a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form. Our results (**Figure** \@ref(fig:PCcorrelation)) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value), which was first reported in Crompton and Gowlett's [-@crompton1993] pioneering study on the allometry of Kilombe handaxes. In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface. Consequently, it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed. It is possible that such constraints help to explain the convergence of our novice knappers on similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove. Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers. This also provides an alternative explanation to the social transmission of form for the experimental convergence of on pointed forms. In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.

@@

```{r comparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Core 63 before (left) and after knapping(right), showing the minimal morphological change during the knapping process."}
knitr::include_graphics(here::here("figure/Fig7.jpg"))
```
